## Choosing a graphic form, Fundamental chart types (Part I)

```{r message = FALSE}
library("chron") # for working with dates / times
library("naniar") # tools for exploring & handing missing data
library("tidytuesdayR") # used to download TidyTuesday data
library("tidycensus") # an R package that allows users to interface with a select number of the US Census Bureau’s data APIs and return tidyverse-ready data frames
library("ggridges") # {ggplot2} extension for creating ridgeline plots
library("gghighlight") # {ggplot2} extension for highlighting geoms
library("ggbeeswarm") # {ggplot2} extension for creating categorical scatter (violin point) plots
library("see") # {ggplot2} extension for model visualization (we'll be using it for it's geom, geom_violindot())
library("scales") # provides the internal scaling infrastructure used by ggplot2, and gives you tools to override the default breaks, labels, transformations and palettes (installed automatically with {ggplot2} or {tidyverse})
```

### Lecture 3.1 Choosing the right graphic form

We understand complex numbers better when they're represented visually

Mapping - how values of a variable of interest are represented by visuals (height of a bar, shaded region of a plot, color of data points)

Q: How do you choose the right graphic form to represent your data?

Different methods

-   length or height

-   poisiton area

-   angle/area

-   hue and shade

Hierarchy of elementary perceptual tasks

-   a viewer performs one or more of these mental-visual tasks to extract the values of real variables represented on most graphs

-   successful cahrts are constructed based on elementary tasks 'as high in the hierarchy as possible'

-   the closer you move to the top of the scale the faster and

Caveats to the hierarchy an example

lower scale methods can be appropriate when the goal is to reveal general patterns. For example, a chloropleth map displays divided geographic areas/regions, which are colored in relation to a numeric numbers

No method of choosing a graphic form is perfect! It's important to think critically about your graphic's purpose and how best to represent your data to serve that purpose

Tips for choosing the right graphic form

-   think about the tasks you want to enable or message you want to convey; do you want to compare, see change or flow, reveal relationships or connections

-   consider the number of variables and the number of data points, as well as the data types you're working with

number of variables and data points, data types

-   continuous: measured data can have an infinite number of values within possible range

    temperature

-   discrete: observations can only exist at limited values, often counts

    -   \# of species counted in a region

nominal - unordered descriptions

ordinal - ordered descriptions

binary - only 2 mutually exclusive outcomes

-   try different graphic forms, especially if you have more than one task to enable or message to convey

If we wanted to show big picture patterns and detailed comparisons, we may consider including multiple graphs

-   arrange the components of the graphic to make it as easy as possible to extract meaning fro your graphic quickly

-   test the outcomes of your graphic on others, particularly on those who are representative of the audience you are trying to reach

Critiquing a data viz:

-   identify primary and secondary insights the graphic is trying to convey

-   identify elementary perceptual tasks and what is confusing or difficult to do

-   identify what's possible to use more effective elementary preceptual tasks for the primary and secondary insights

-   identify points of confusion

### Lecture 3.2 Visualizing Data Distributions

Exploring bottom temperatures recorded at Mohawk Reef, a near-shore rocky reef and one of the Santa Barbara Coastal LTER research sites

```{r message = false}
# load packages
library(tidyverse)
library(chron)
library(naniar)

# import data ---
mko <- read_csv('https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.2007.17&entityid=02629ecc08a536972dec021f662428aa')
```

#### Data Wrangling

```{r, warning=FALSE}
mko_clean <- mko %>% 
  select(year, month, day, decimal_time, Temp_bot) %>% 
  unite(date, year, month, day, sep = '-',remove = FALSE) %>% 
  mutate(time = times(decimal_time)) %>% 
  unite(date_time, date, time, sep = ' ') %>% 
  # treat it as a date time object; overwrite date time column
  mutate(date_time = as.POSIXct(date_time, "%Y-%m-%d %H:%M:%S", tx = 'GMZ'),
         year = as.factor(year), #treating ymd as a categorical variable
         month = as.factor(month),
         day = as.numeric(day)) %>% 
  mutate(month_name = as.factor(month.name[month])) %>% 
  replace_with_na(replace = list(Temp_bot = 9999))
  
  

```

##### Histograns - `ggplot::geom_histogram()`

-   important considerations : bin width (30 bins by default); too few/too many bins

```{r}
# === original plot ===
# position geom_histogram uses is stacked 
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name))+
  geom_histogram(position = 'identity', alpha = 0.5)


# small multiples --
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  ggplot(aes(x = Temp_bot))+
  geom_histogram(position = 'identity', alpha = 0.5)+
  facet_wrap(~month_name)

# fewer groups at a time --- 
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  filter(month_name %in% c('April','June','October')) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name))+
  geom_histogram(position = 'identity', alpha = 0.5, bins = 1) + 
  scale_fill_manual(values = c("#2C5374","#ADD8E6","#8B3A3A"))
```

#### Density plots - `ggplot2::geom_density()`

A smoothed version of a histogram. Density plots are representations of the distribution of numeric variables; uses a KDE to show the probability density function of a variable. The area under each curve is equal to 1. Use a density plot when you are most concerned with the shape of the distribution.

```{r}
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name))+
  geom_density( alpha = 0.5)


# fewer groups at a time --- 
mko_clean %>% 
  mutate(month_name = factor(month_name, levels = month.name)) %>% 
  filter(month_name %in% c('April','June','October')) %>% 
  ggplot(aes(x = Temp_bot, fill = month_name))+
  geom_density(alpha = 0.5, adjust = 1/2) + 
  scale_fill_manual(values = c("#2C5374","#ADD8E6","#8B3A3A"))
```

Histograms show us the counts of values in each range, represented by the height of the bars

density plots show the proportion of values in each area (area under curve equal 1; peaks indicate where more values are concentrated, but it does not tell us anything about the number of observations)

##### Hist vs density

```{r}
dummy_data <- data.frame(value = c(rnorm(n = 100, mean = 5),
                                   rnorm(n = 200, mean = 10)),
                         group = rep(c("A", "B"),
                                     times = c(100, 200)))

# histogram --
ggplot(dummy_data, aes(x = value, fill = group))+
  geom_histogram(position = 'identity', alpha = 0.7)+
  geom_rug(aes(color = group), alpha = 0.7)

# density --
ggplot(dummy_data, aes(x = value, fill = group))+
  geom_density(alpha = 0.7)+
  geom_rug(aes(color = group), alpha = 0.7)
```

#### Ridgeline Plots - `ggridges`

they show the distribution of a numeric variable for multiple groups

need : a numeric variable with lots of values

important considerations - work best when you have more than 6 groups

```{r}
ggplot(mko_clean, aes(x = Temp_bot, y = month_name, 
                      fill = after_stat(x)))+
  ggridges::geom_density_ridges_gradient(rel_min_height = 0.01, scale = 3)+
  scale_y_discrete(limits = rev(month.name))+
  scale_fill_gradientn(colors = c("#2C5374","#849BB4", "#D9E7EC", "#EF8080", "#8B3A3A"))
```

adjustments: jitter raw data; visualize the raw data underlying the density ridges ( since our temperature data is too large)

#### Box plots - `ggplot2::geom_boxplot()`

what are they?

-   box plots summarize the distribution of a numeric variable for one or several groups

need: a numeric variable, often with multiple groups

```{r}
ggplot(mko_clean, aes(x = month_name, y = Temp_bot, fill = month_name))+
  geom_boxplot()+
  scale_x_discrete(limits = rev(month.name)) + 
  coord_flip()+
  gghighlight::gghighlight(month_name == 'October')+
  theme(
    legend.position = 'none'
  )

# jitter and dodge 
library(palmerpenguins)

penguins %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot(aes(x = species, y = body_mass_g, color = year))+
  geom_boxplot()+
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.2))+
  coord_flip()
```

#### Violin plots - `ggplot2::geom_violin()`

-   visualize the distribution of a numeric variable for one or several groups, where the shape of the violin represents the density estimate of the variable

```{r}
ggplot(mko_clean, aes(x = month_name, y = Temp_bot))+
  geom_violin()+
  geom_boxplot(color = 'gray', alpha = 0.5, width = 0.1, outlier.color = 'black')+
  scale_x_discrete(limits = rev(month.name))+
  coord_flip()


# half n half 
ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species))+
  see::geom_violindot(size_dots = 5, alpha = 0.5)
```

### Lecture 3.3 Tidycensus

```{r message = FALSE}
library(tidyverse)
library(tidycensus)
library(janitor)
library(gghighlight)

# use the source function to execute everything inthat file

source(here::here('week-3','KEYS.R'))
census_api_key(censusKEY)

lyme <- read_csv(here::here('week-3','data','LD-Case-Counts-by-County-01-20.csv'))
```

### Wrangle

```{r}
# -- lyme data 
lyme_clean <- lyme %>% 
  clean_names() %>% 
  rename(county = ctyname,
         state = stname,
         status = ststatus) %>% 
  pivot_longer(cols = 6:25,
               names_to = 'city_year',
               values_to = 'reported_cases') %>% 
  mutate(year = str_remove(city_year, pattern = 'cases'),
         year = as.factor(year))

# calculate the total cases per state 
lyme_by_state <- lyme_clean %>% 
  group_by(year, state) %>% 
  summarize(total_cases = sum(reported_cases))

us_state_pop <- get_estimates(geography = 'state',
                              product = 'population',
                              year = 2019) %>% 
  filter(variable == 'POP') %>% 
  select(state = NAME, 
         population = value)

lyme_pop <- left_join(lyme_by_state, us_state_pop) %>% 
  mutate(pop100k = population/100000) %>% 
  mutate(cases_per_100k = total_cases/pop100k)

```

## Line Plot

```{r}
# highlight 1 group -----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state, color = state))+
  geom_line()+
  gghighlight::gghighlight(state == 'New Jersey')

# highlight based on a conditional -------
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state, color = state))+
  geom_line()+
  gghighlight::gghighlight(max(cases_per_100k) > 100)
```

Do not start at the y-axis at 0 if the range of data is small but the distance from the bottom of the range to 0 is large.

The larger the aspect ratio, the steeper changes appear, which may cause readers to interpret changes as more important. The smaller the aspect ratio, the flatter the line which may cause readers to interpret changes as small/insignificant

Aspect ratio affects perception of slope ; with a wider aspect ratio, we can more clearly see the differences in rates of change in sunspot activity

```{r}
# aspect ratio ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state == 'Vermont') %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state))+
  geom_line()+
  scale_y_continuous(breaks = seq(0,190, by = 1))+
  coord_fixed(ratio = 1/20)
```

### Area plots

```{r}
# area chart 1 group ----
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state == "Washington") %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state, fill = state))+
  geom_area()+
  scale_fill_manual(values = c("blue"))+
  scale_x_discrete(expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))+
  theme(legend.position = 'none')


# stacked area chart ------
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state %in% c("Maine", "Rhode Island","New Hampshire","Vermont")) %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state, fill = state))+
  geom_area(position = 'fill', rev =)+
  scale_y_continuous(labels = scales::label_percent(scale = 100))


# line chart version
lyme_pop %>% 
  filter(year %in% c(2010:2020)) %>% 
  filter(state %in% c("Maine", "Rhode Island","New Hampshire","Vermont")) %>% 
  ggplot(aes(x = year, y = cases_per_100k, group = state, fill = state))+
  geom_line()+
  facet_wrap(~state)
```
